Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity

### 2021-12-02 paper.tex
Fix reviewers language comments

In addition to our main goal, of particular interest here is also the effect of
anisotropic plasma viscosity, which in the following is found to strongly
influence the growth of the flute instability. It is well known that viscosity
in magnetised plasmas (such as those which make up the solar corona) is
anisotropic and strongly dependent on the strength and direction of the local
magnetic
field~\citep{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
To take this into account, MacTaggaret et al. developed a phenomenological
model of anisotropic
viscosity~\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017} that
captures the main physics of viscosity in the solar corona as outlined in the
analysis of~\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity in
regions of very weak or zero field strength. For brevity, we will refer to this
model of viscosity as ``the switching model''.
In~\citet{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}
we implemented the switching model as a module for the widely-used general MHD
code Lare3d~\citep{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the development of the
nonlinear MHD kink instability and the Kelvin-Helmholtz instability. More
generally, the interest in anisotropic viscosity stems from the open question
of which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\citep{klimchukSolvingCoronalHeating2006}, an important facet of solving
the coronal heating problem. Using scaling laws, it has been suggested that
viscous heating (generated through anisotropic viscosity) can dwarf that of
Ohmic
heating~\citep{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However, due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small scale
instabilities and turbulence is relatively
unknown~\citet{klimchukSolvingCoronalHeating2006}. In addition to directly
heating the plasma, viscosity plays a part in the damping of instabilities and
waves~\citep{rudermanSlowSurfaceWave2000}. It is this effect we are most
interested in here, and it shall be reported that the use of anisotropic
viscosity permits the growth of the flute instability, which is otherwise
strongly damped by isotropic viscosity.

@@

Throughout this paper, the twisted flux tube generated by the drivers has
a pressure profile which is approximately axisymmetric, and independent of $z$
away from the boundaries at $z=\pm2$, and has a negative gradient, hence
$|\nabla p|$ may be written as $-d p/ dr$. Similarly, away from the boundaries,
the magnetic field has a negligible $r$ component and little dependence on
$\theta$ and $z$, allowing the field to be approximated as $\vec{B} = (0,
B_{\theta}(r), B_z(r))^{\text{T}}$, in cylindrical coordinates $(r, \theta,
z)$. For a twisted field of this form, the radius of curvature is given by
\begin{equation} \label{eq:radius_of_curvature} R_c
= \frac{1}{|(\vec{b}\cdot\nabla) \vec{b}|} = \frac{r}{b_{\theta}^2},
\end{equation} where $\vec{b} = \vec{B}/|\vec{B}|$ is the unit vector in the
direction of the magnetic field and $b_{\theta}$ is the component of $\vec{b}$
in the azimuthal direction. These approximations allows the growth rate to be
written as
\begin{equation}
\label{eq:fluting_growth_rate2}
\gamma_{\text{ideal}}^2 = \frac{-2p'}{\rho R_c}.
\end{equation}
This approximation for the growth rate continues to hold while the flux tube
remains relatively axisymmetric, that is while the kink instability remains in
its linear phase.

@@

In all cases unstable to the flute instability, it occurs  between $t=20$ and
$t=30$. The continued driving at the boundaries eventually injects enough twist
that the tube also becomes unstable to the kink instability. This initially
develops linearly alongside or shortly after the flute instability and then
erupts during the its  nonlinear phase, dominating the dynamics and disrupting
the   flute instability. The onset and the competition of the two instabilities
is strongly affected by the value of $\eta$ and the viscosity model used.

@@

We now describe the evolution and competition of flute and kink instabilities
in case of resistivity $\eta=10^{-4}$. Figure~\ref{fig:kink_pressure_slices-4}
shows the pressure profile of the anisotropic viscosity case
\eqref{eq:switching_model} at time moments $t=26$, $28$ and $30$ and at $z=0$.
At $t=26$ the tube becomes unstable to  flute instability with azimuthal
wavenumber $m=4$, when the plasma begins to  bulge out diagonally from the
high-pressure core as seen in figure~\ref{fig:swi-4_pressure_13}. As  the
bulges move radially outwards into lower pressure regions they expand and
accelerate, resulting in the entire pressure structure appearing taking the
shape of a four-leaf clover (figure~\ref{fig:swi-4_pressure_14}). By $t=30$ the
kink instability has disrupted the flute instability and is developing
nonlinearly as evident in figure~\ref{fig:swi-4_pressure_15}. As is typical of
nonlinear kink development, the tube continues to release its stored potential
energy in the form of kinetic energy and heat and the contained plasma becomes
highly mixed. In the  isotropic viscosity case which will not be illustrated by
a separate figure, the flute instability is present but damped, and it is
quickly outcompeted by the kink instability which quickly dominates the
dynamics.

@@

Figure~\ref{fig:kink_field_line_plots} shows the effect the viscosity models
have on the initial stages of the flute and kink instabilities in 3D. While the
flute instability is observed in both cases, it is damped in the isotropic case
and grows faster in the anisotropic case. In the latter case, the extended
development of the flute instability appears to disrupt the inner core of field
lines and, as will be discussed further below, slows the growth of the kink
instability. In the isotropic case, the flute instability has been damped to
the extent that the kink instability grows uninhibited and quickly disrupts the
fluting.

@@

Figure \ref{fig:kink_pressure_slices-3} shows a prolonged development of the
flute instability and a slow nonlinear development of the kink instability at
the higher resistivity value $\eta=10^{-3}$ in the case of anisotropic
viscosity. Due to the enhanced Ohmic heating at $\eta=10^{-3}$, the pressure
gradient is substantially stronger than at $\eta=10^{-4}$ and the flute
instability is excited much earlier. Compared to the $\eta=10^{-4}$ cases, the
instability occurs further from the axis, at $r\approx0.16$, and the larger
pressure gradient drives the bulges in profile further from the axis during the
nonlinear phase (figure~\ref{fig:swi-3_pressure_12}). These bulges continue to
extend outwards and mix the plasma as they develop. The kink instability can be
observed moving the axis of the tube diagonally upwards and to the right in
figure~\ref{fig:swi-3_pressure_15}. At this time in the $\eta=10^{-4}$ cases,
the nonlinear development of the kink was at a later stage of its development
(figure~\ref{fig:swi-4_pressure_15}). The development of the kink then proceeds
slowly as it moves the axis of the tube through the mixed region to eventually
begin the reconnection process with the outer region of field that is typical
of the instability in this kind of flux tube (as was observed in our earlier
work~\citep{quinnEffectAnisotropicViscosity2020}).

>>>>

In addition to our main goal, of particular interest here is the effect of
anisotropic plasma viscosity, which in the following is found to strongly
influence the growth of the flute instability. It is well known that viscosity
in magnetised plasmas (such as those which make up the solar corona) is
anisotropic and strongly dependent on the strength and direction of the local
magnetic
field~\citep{hollwegViscosityChewGoldbergerLowEquations1986,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
To take this into account, MacTaggaret et al. developed a phenomenological
model of anisotropic
viscosity~\citep{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017} that
captures the main physics of viscosity in the solar corona as outlined in the
analysis of~\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity in
regions of very weak or zero field strength. For brevity, we will refer to this
model of viscosity as ``the switching model''.
In~\citet{quinnEffectAnisotropicViscosity2020,quinnKelvinHelmholtzInstabilityCollapse2021}
we implemented the switching model as a module for the widely-used general MHD
code Lare3d~\citep{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the development of the
nonlinear MHD kink instability and the Kelvin-Helmholtz instability. More
generally, the interest in anisotropic viscosity stems from the open question
of which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\citep{klimchukSolvingCoronalHeating2006}, an important facet of solving
the coronal heating problem. Using scaling laws, it has been suggested that
viscous heating (generated through anisotropic viscosity) can dwarf that of
Ohmic
heating~\citep{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However, due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small scale
instabilities and turbulence is relatively
unknown~\citet{klimchukSolvingCoronalHeating2006}. In addition to directly
heating the plasma, viscosity plays a part in the damping of instabilities and
waves~\citep{rudermanSlowSurfaceWave2000}. It is this effect we are most
interested in here, and it shall be reported that the use of anisotropic
viscosity permits the growth of the flute instability, which is otherwise
strongly damped by isotropic viscosity.

@@

Throughout this paper, the twisted flux tube generated by the drivers has
a pressure profile which is approximately axisymmetric, and independent of $z$
away from the boundaries at $z=\pm2$, and has a negative gradient, hence
$|\nabla p|$ may be written as $-d p/ dr$. Similarly, away from the boundaries,
the magnetic field has a negligible $r$ component and little dependence on
$\theta$ and $z$, allowing the field to be approximated as $\vec{B} = (0,
B_{\theta}(r), B_z(r))^{\text{T}}$, in cylindrical coordinates $(r, \theta,
z)$. For a twisted field of this form, the radius of curvature is given by
\begin{equation} \label{eq:radius_of_curvature} R_c
= \frac{1}{|(\vec{b}\cdot\nabla) \vec{b}|} = \frac{r}{b_{\theta}^2},
\end{equation} where $\vec{b} = \vec{B}/|\vec{B}|$ is the unit vector in the
direction of the magnetic field and $b_{\theta}$ is the component of $\vec{b}$
in the azimuthal direction. These approximations allow the growth rate to be
written as
\begin{equation}
\label{eq:fluting_growth_rate2}
\gamma_{\text{ideal}}^2 = \frac{-2p'}{\rho R_c}.
\end{equation}
This approximation for the growth rate continues to hold while the flux tube
remains relatively axisymmetric, that is while the kink instability remains in
its linear phase.

@@

In all cases unstable to the flute instability, it occurs  between $t=20$ and
$t=30$. The continued driving at the boundaries eventually injects enough twist
that the tube also becomes unstable to the kink instability. This initially
develops linearly alongside or shortly after the flute instability and then
erupts during its nonlinear phase, dominating the dynamics and disrupting
the   flute instability. The onset and the competition of the two instabilities
is strongly affected by the value of $\eta$ and the viscosity model used.

@@

We now describe the evolution and competition of flute and kink instabilities
in case of resistivity $\eta=10^{-4}$. Figure~\ref{fig:kink_pressure_slices-4}
shows the pressure profile of the anisotropic viscosity case
\eqref{eq:switching_model} at time moments $t=26$, $28$ and $30$ and at $z=0$.
At $t=26$ the tube becomes unstable to the flute instability with azimuthal
wavenumber $m=4$, when the plasma begins to  bulge out diagonally from the
high-pressure core as seen in figure~\ref{fig:swi-4_pressure_13}. As  the
bulges move radially outwards into lower pressure regions they expand and
accelerate, resulting in the entire pressure structure appearing taking the
shape of a four-leaf clover (figure~\ref{fig:swi-4_pressure_14}). By $t=30$ the
kink instability has disrupted the flute instability and is developing
nonlinearly as evident in figure~\ref{fig:swi-4_pressure_15}. As is typical of
nonlinear kink development, the tube continues to release its stored potential
energy in the form of kinetic energy and heat and the contained plasma becomes
highly mixed. In the  isotropic viscosity case which will not be illustrated by
a separate figure, the flute instability is present but damped, and it is
quickly outcompeted by the kink instability which quickly dominates the
dynamics.

@@

Figure~\ref{fig:kink_field_line_plots} shows the effect the viscosity models
have on the initial stages of the flute and kink instabilities in 3D. While the
flute instability is observed in both cases, it is damped in the isotropic case
and grows faster in the anisotropic case. In the latter case, the extended
development of the flute instability appears to disrupt the inner core of field
lines and, as will be discussed further below, slows the growth of the kink
instability. In the isotropic case the flute instability has been damped to
the extent that the kink instability grows uninhibited and quickly disrupts the
fluting.

@@

Figure \ref{fig:kink_pressure_slices-3} shows a prolonged development of the
flute instability and a slow nonlinear development of the kink instability at
the higher resistivity value $\eta=10^{-3}$ in the case of anisotropic
viscosity. Due to the enhanced Ohmic heating at $\eta=10^{-3}$, the pressure
gradient is substantially stronger than at $\eta=10^{-4}$ and the flute
instability is excited much earlier. Compared to the $\eta=10^{-4}$ cases, the
instability occurs further from the axis, at $r\approx0.16$, and the larger
pressure gradient drives the bulges in profile further from the axis during the
nonlinear phase (figure~\ref{fig:swi-3_pressure_12}). These bulges continue to
extend outwards and mix the plasma as they develop. The kink instability can be
observed moving the axis of the tube diagonally upwards and to the right in
figure~\ref{fig:swi-3_pressure_15}. At this time in the $\eta=10^{-4}$ cases,
the nonlinear development of the kink was at a later stage of its development
(figure~\ref{fig:swi-4_pressure_15}). The development of the kink then proceeds
slowly as it moves the axis of the tube through the mixed region to eventually
begin the reconnection process with the outer region of field that is typical
of the instability in this kind of flux tube (as was observed in our earlier
work~\citep[][]{quinnEffectAnisotropicViscosity2020}).
